---
title: "Hands-on Exercise 1-1: Geospatial Data Wrangling with R"
editor: visual
---

# Chapter 1: Geospatial Data Wrangling with R

## 1 Overview

In this hands-on exercise, I learn how to import and wrangling geospatial data using appropriate R packages.

## 2 Getting Started

The code chunk below installs and loads [sf](https://r-spatial.github.io/sf/) and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## 3 Importing Geospatial Data

### 3.1 Importing polygon feature data

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### 3.2 Importing polyline feature data in shapefile form

```{r}
cyclingpath <- st_read(dsn = "data/geospatial", layer = "CyclingPath")
```

### 3.3 Importing GIS data in kml format

```{r}
preschool <- st_read("data/geospatial/pre-schools-location-kml.kml")
```

## 4 Checking the Content of A Simple Feature Data Frame

### 4.1 Working with *st_geometry()*

```{r}
st_geometry(mpsz)
```

### 4.2 Working with glimpse()

```{r}
glimpse(mpsz)
```

### 4.3 Working with head()

```{r}
head(mpsz, n=5)
```

## 5 Plotting the Geospatial Data

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

## 6 Working with Projection

### 6.1 Assigning EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

```{r}
mpsz3414 <- st_set_crs(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

### 6.2 Transforming the projection of preschool from wgs84 to svy21.

```{r}
preschool3414 <- st_transform(preschool, crs = 3414)
```

## 7 Importing and Converting An Aspatial Data

### 7.1 Importing the aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
list(listings)
```

### 7.2 Creating a simple feature data frame from an aspatial data frame

```{r}
listings_sf <- st_as_sf(listings, coords = c("longitude","latitude"), crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

## 8 Geoprocessing with sf package

### 8.1 Buffering

```{r}
buffer_cycling <- st_buffer(cyclingpath, dist = 5, nQuadSegs = 30)
```

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

```{r}
sum(buffer_cycling$AREA)
```

### 8.2 Point-in-polygon count

```{r}
mpsz3414$`PreSch Count` <- lengths(st_intersects(mpsz3414, preschool3414))
```

```{r}
summary(mpsz3414$`PreSch Count`)
```

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
```

```{r}
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

## 9 Exploratory Data Analysis (EDA)

```{r}
hist(mpsz3414$`PreSch Density`)
```

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

```{r}
ggplot(mpsz3414, 
       aes(x=as.numeric(`PreSch Density`), y=as.numeric(`PreSch Count`)))+
  geom_point() +
  labs(title = "Scatterplot: Pre-school density vs Pre-school count",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```
